---
title: "CellHet Package Examples"
author: "CellHet Team"
date: "April 2025"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 7
)
library(Seurat)
library(ggplot2)
library(dplyr)
library(patchwork)
```

# Introduction to CellHet

CellHet is a comprehensive R package for analyzing cellular heterogeneity in single-cell RNA-seq data across multiple conditions, cell types, and data modalities. This document provides practical examples of how to use CellHet's core functionality.

## Installation

```{r installation, eval=FALSE}
# Install CellHet from CRAN
install.packages("CellHet")

# Or install the development version from GitHub
# install.packages("devtools")
# devtools::install_github("guokai8/CellHet")
```

```{r load-library}
library(CellHet)
```

# Basic Example: DEG Analysis

First, let's demonstrate how to perform differential expression analysis across cell types and conditions.

## Load and preprocess data

For this example, we'll use a PBMC dataset from Seurat.

```{r load-data}
# Load example dataset
data("pbmc_small")

# Add a simulated condition variable for demonstration
set.seed(42)
pbmc_small$condition <- sample(c("Control", "Treatment"), ncol(pbmc_small), replace = TRUE)

# Process the data
pbmc_small <- NormalizeData(pbmc_small)
pbmc_small <- FindVariableFeatures(pbmc_small)
pbmc_small <- ScaleData(pbmc_small)
pbmc_small <- RunPCA(pbmc_small)
pbmc_small <- FindNeighbors(pbmc_small)
pbmc_small <- FindClusters(pbmc_small, resolution = 0.5)
pbmc_small <- RunUMAP(pbmc_small, dims = 1:10)

# Visualize UMAP with clusters
DimPlot(pbmc_small, group.by = "group", label = TRUE) + 
  ggtitle("PBMC Clusters")
```

## Run Differential Expression Analysis

Now, let's use CellHet to find differentially expressed genes between conditions across all cell types:

```{r run-deg-analysis}
# Identify differentially expressed genes
deg_results <- findDifferentialGenes(
  object = pbmc_small,
  group_var = "group",
  cell_type_var = "seurat_clusters",
  min_cells = 3,
  logfc.threshold = 0.25,
  p_val_adj_threshold = 0.05,
  test.use = "wilcox",reference_group = "HFD",
  return_summary = TRUE
)

# View summary results
head(deg_results$summary)
```

## Visualize DEG Results

CellHet provides several functions for visualizing DEG results:

```{r visualize-degs}
# Create a heatmap of DEGs across cell types
heatmap_plot <- visualizeDEGHeatmap(
  deg_results = deg_results,
  show_reg_indicators = TRUE,
  top_labels = TRUE,
  cluster_rows = TRUE
)
print(heatmap_plot)

# Create a barplot of DEGs
barplot <- visualizeDEGBarplot(
  deg_results = deg_results,
  up_color = "darkorange",
  down_color = "deepskyblue",
  facet_by = "comparison"
)
print(barplot)

# Create a dotplot of DEGs
dotplot <- visualizeDEGDotplot(
  deg_results = deg_results,
  count_display = "inside",
  point_size_range = c(1, 10)
)
print(dotplot)
```

## Find Shared DEGs Across Cell Types

You can identify DEGs that are shared across multiple cell types:

```{r shared-degs}
# Find shared DEGs
shared_degs <- findSharedDEGs(
  deg_results = deg_results,
  min_deg_count = 1,
  plot_type = "upset",
  direction = "both"
)

# Visualize shared DEGs
print(shared_degs$plot)
```

## Compare Cell Type Responses

You can also compare how different cell types respond to treatment:

```{r compare-responses}
# Compare responses across cell types
response_comparison <- compareResponses(
  deg_results = deg_results,
  reference_cell_type = "0",
  plot_type = "heatmap",
  similarity_metric = "jaccard"
)

# Visualize response comparison
print(response_comparison$plot)
```

# Advanced Example: Trajectory Analysis

CellHet includes functions for analyzing gene expression changes along a trajectory.

```{r trajectory-setup, eval=FALSE}
# For this example, we assume a pseudotime has been calculated
# In practice, you would use tools like Monocle3, Slingshot, or Velocyto
# This is just a simulated pseudotime for demonstration

# Add a simulated pseudotime variable
set.seed(42)
pbmc_small$pseudotime <- runif(ncol(pbmc_small))

# Run trajectory analysis
trajectory_results <- analyzeDEGTrajectory(
  sce = pbmc_small,
  trajectory_var = "pseudotime",
  cell_type_var = "seurat_clusters",
  n_bins = 10,
  smooth_method = "loess",
  n_top_genes = 30,
  cluster_genes = TRUE,
  n_clusters = 3,
  plot_type = "both"
)

# Visualize gene expression along the trajectory
trajectory_plot <- plotGeneTrajectory(
  trajectory_results = trajectory_results,
  genes = c("MS4A1", "CD79A", "LYZ"),
  plot_type = "line",
  add_loess_fit = TRUE
)

print(trajectory_plot)

# Compare trajectories between conditions
diff_trajectory <- diffTrajectoryAnalysis(
  sce = pbmc_small,
  trajectory_var = "pseudotime",
  group_var = "condition",
  cell_type_var = "seurat_clusters",
  n_bins = 10,
  test_method = "anova",
  p_val_threshold = 0.05
)

# View top differentially expressed genes along trajectory
head(diff_trajectory$test_results)
```

# Advanced Example: Spatial Analysis

CellHet provides tools for analyzing spatial heterogeneity in spatial transcriptomics data.

```{r spatial-analysis, eval=FALSE}
# For this example, we would use a Seurat object with spatial data
# or a SpatialExperiment object
# This is just pseudo-code for demonstration

# Analyze spatial heterogeneity
spatial_results <- analyzeSpatialHeterogeneity(
  spatial_object = spatial_object,
  cell_type_col = "cell_type",
  spatial_coords = c("x", "y"),
  method = "morans_i",
  cluster_method = "dbscan",
  n_clusters = 5
)

# Create a spatial heterogeneity map
heterogeneity_map <- plotSpatialHeterogeneity(
  spatial_results = spatial_results,
  plot_type = "all",
  top_n = 5,
  color_palette = "viridis",
  add_convex_hull = TRUE
)

print(heterogeneity_map)

# Analyze spatial co-occurrence of cell types
cooccurrence_results <- analyzeSpatialCoOccurrence(
  spatial_object = spatial_object,
  cell_type_col = "cell_type",
  spatial_coords = c("x", "y"),
  radius = 50,
  method = "correlation",
  plot_type = "both"
)

print(cooccurrence_results$plots$heatmap)
print(cooccurrence_results$plots$network)
```

# Advanced Example: Multi-Modal Integration

CellHet can integrate and analyze multiple data modalities such as scRNA-seq, scATAC-seq, and spatial transcriptomics.

```{r multi-modal, eval=FALSE}
# For this example, we would use multiple Seurat or SingleCellExperiment objects
# This is just pseudo-code for demonstration

# Integrate multiple modalities
integration_results <- integrateMultiModalData(
  data_list = list(rna_data, atac_data),
  modality_names = c("RNA", "RNA2"),
  cell_type_var = "cell_type",
  integration_method = "harmony",
  dim_reduction = "umap"
)

# Visualize integrated data
print(integration_results$plots$modality_plot)
print(integration_results$plots$celltype_plot)

# Analyze heterogeneity across modalities
heterogeneity_results <- analyzeMultiModalHeterogeneity(
  integration_results = integration_results,
  n_markers = 20,
  test_method = "wilcox",
  plot_heatmap = TRUE
)

# Compare responses across modalities
response_comparison <- compareMultiModalResponses(
  data_list = list(rna_data, atac_data),
  modality_names = c("RNA", "ATAC"),
  cell_type_var = "cell_type",
  condition_var = "condition",
  reference_condition = "Control",
  plot_type = "both"
)

# View shared differential features across modalities
print(response_comparison$plots$summary_barplot)
```

# Advanced Use Case: Single-cell Multi-omic Analysis

```{r multi-omic-example, eval=FALSE}
# Example of integrating RNA-seq and ATAC-seq data from the same cells
# This would use real multi-omic data in practice

# Create a custom visualized workflow
multiOmicWorkflow <- function(rna_data, atac_data, cell_metadata) {
  # Preprocess RNA data
  rna_processed <- findDifferentialGenes(
    object = rna_data,
    group_var = "condition",
    cell_type_var = "cell_type"
  )
  
  # Preprocess ATAC data
  atac_processed <- findDifferentialGenes(
    object = atac_data,
    group_var = "condition",
    cell_type_var = "cell_type"
  )
  
  # Integrate modalities
  integrated_data <- integrateMultiModalData(
    data_list = list(rna_data, atac_data),
    modality_names = c("RNA", "ATAC"),
    cell_type_var = "cell_type"
  )
  
  # Analyze cell type heterogeneity across modalities
  heterogeneity_analysis <- analyzeMultiModalHeterogeneity(
    integration_results = integrated_data
  )
  
  # Identify conserved and modality-specific markers
  conserved_markers <- heterogeneity_analysis$conserved_markers
  rna_specific <- heterogeneity_analysis$modality_specific_markers$RNA
  atac_specific <- heterogeneity_analysis$modality_specific_markers$ATAC
  
  # Generate visualizations
  plots <- list(
    integrated_umap = integrated_data$plots$celltype_plot,
    heatmap_conserved = heterogeneity_analysis$plots$conserved_T_cells,
    heatmap_rna = heterogeneity_analysis$plots$RNA_B_cells
  )
  
  return(list(
    integrated_data = integrated_data,
    markers = list(
      conserved = conserved_markers,
      rna_specific = rna_specific,
      atac_specific = atac_specific
    ),
    plots = plots
  ))
}

# Example usage:
# results <- multiOmicWorkflow(rna_data, atac_data, cell_metadata)
# print(results$plots$integrated_umap)
```

# Combining CellHet with Other Tools

CellHet works well with other popular single-cell analysis packages. Here's an example of combining CellHet with other tools:

```{r combined-workflow, eval=FALSE}
# Example workflow combining Seurat, scran, and CellHet

# 1. Initial preprocessing with Seurat
seurat_obj <- CreateSeuratObject(counts = raw_counts)
seurat_obj <- NormalizeData(seurat_obj)
seurat_obj <- FindVariableFeatures(seurat_obj)
seurat_obj <- ScaleData(seurat_obj)
seurat_obj <- RunPCA(seurat_obj)
seurat_obj <- FindNeighbors(seurat_obj)
seurat_obj <- FindClusters(seurat_obj)
seurat_obj <- RunUMAP(seurat_obj, dims = 1:30)

# 2. Transfer to SingleCellExperiment for additional analyses
library(scater)
library(scran)
sce <- as.SingleCellExperiment(seurat_obj)
sce <- computeSumFactors(sce)
sce <- logNormCounts(sce)
var_genes <- getTopHVGs(sce, n = 2000)

# 3. Back to Seurat for trajectory analysis
seurat_obj <- FindVariableFeatures(seurat_obj, selection.method = "vst", nfeatures = 2000)
seurat_obj <- RunVelocity(seurat_obj) # Using SeuratWrappers
seurat_obj$pseudotime <- calculatePseudotime(seurat_obj) # Pseudocode function

# 4. CellHet for advanced analysis
# Differential expression analysis
deg_results <- findDifferentialGenes(
  object = seurat_obj,
  group_var = "condition",
  cell_type_var = "seurat_clusters"
)

# Cell type heterogeneity analysis
heterogeneity_map <- visualizeHeterogeneity(
  deg_results = deg_results,
  layout_type = "network",
  level = "gene"
)

# Trajectory analysis
trajectory_results <- analyzeDEGTrajectory(
  sce = seurat_obj,
  trajectory_var = "pseudotime",
  cell_type_var = "seurat_clusters"
)

# Integrate results
all_results <- list(
  degs = deg_results,
  heterogeneity = heterogeneity_map,
  trajectory = trajectory_results
)

# Visualize combined results
trajectory_degs <- intersect(
  trajectory_results$top_genes,
  rownames(deg_results$all_degs[deg_results$all_degs$significant, ])
)

# Plot trajectory for DEGs
gene_trajectory_plot <- plotGeneTrajectory(
  trajectory_results = trajectory_results,
  genes = trajectory_degs[1:5]
)

print(gene_trajectory_plot)
```

# Visualizing Results with Interactive Shiny Apps

CellHet outputs can be easily integrated into interactive Shiny applications for better data exploration:

```{r shiny-app, eval=FALSE}
# Create a simple Shiny app for DEG exploration
library(shiny)
library(DT)

createDEGExplorer <- function(deg_results) {
  # Define UI
  ui <- fluidPage(
    titlePanel("CellHet DEG Explorer"),
    sidebarLayout(
      sidebarPanel(
        selectInput("cellType", "Cell Type:",
                    choices = unique(deg_results$summary$cell_type)),
        selectInput("comparison", "Comparison:",
                    choices = unique(deg_results$summary$comparison)),
        sliderInput("pvalCutoff", "P-value Cutoff:",
                    min = 0, max = 0.1, value = 0.05, step = 0.01),
        radioButtons("direction", "Direction:",
                    choices = c("Both" = "both", "Up" = "up", "Down" = "down"),
                    selected = "both")
      ),
      mainPanel(
        tabsetPanel(
          tabPanel("DEG Table", DTOutput("degTable")),
          tabPanel("Visualizations", 
                   plotOutput("heatmapPlot"),
                   plotOutput("barPlot"))
        )
      )
    )
  )
  
  # Define server
  server <- function(input, output, session) {
    # Reactive filtered data
    filteredDEGs <- reactive({
      # Get result key
      result_key <- paste0(input$cellType, "__", input$comparison)
      
      # Get DEGs for this key
      if (!result_key %in% names(deg_results$degs)) {
        return(data.frame())
      }
      
      degs <- deg_results$degs[[result_key]]
      
      # Filter by p-value
      degs <- degs[degs$p_val_adj < input$pvalCutoff, ]
      
      # Filter by direction
      if (input$direction == "up") {
        degs <- degs[degs$avg_log2FC > 0, ]
      } else if (input$direction == "down") {
        degs <- degs[degs$avg_log2FC < 0, ]
      }
      
      return(degs)
    })
    
    # Render DEG table
    output$degTable <- renderDT({
      degs <- filteredDEGs()
      if (nrow(degs) == 0) {
        return(data.frame(Message = "No DEGs match the criteria"))
      }
      
      # Format table
      degs <- degs[order(degs$p_val_adj), ]
      degs$p_val <- formatC(degs$p_val, format = "e", digits = 2)
      degs$p_val_adj <- formatC(degs$p_val_adj, format = "e", digits = 2)
      degs$avg_log2FC <- round(degs$avg_log2FC, 3)
      
      return(degs)
    })
    
    # Render heatmap plot
    output$heatmapPlot <- renderPlot({
      visualizeDEGHeatmap(
        deg_results = deg_results,
        show_reg_indicators = TRUE
      )
    })
    
    # Render barplot
    output$barPlot <- renderPlot({
      visualizeDEGBarplot(
        deg_results = deg_results,
        facet_by = "comparison"
      )
    })
  }
  
  # Create and return the Shiny app
  shinyApp(ui, server)
}

# Example usage:
# app <- createDEGExplorer(deg_results)
# runApp(app)
```

# Working with Large Datasets

CellHet is designed to work efficiently with large single-cell datasets. Here are some tips:

```{r large-datasets, eval=FALSE}
# For large datasets, consider using these settings:

# 1. Use parallel processing
deg_results <- findDifferentialGenes(
  object = large_seurat_obj,
  group_var = "condition",
  cell_type_var = "cell_type",
  workers = 8,  # Use 8 cores for parallel processing
  future.globals.maxSize = 4 * 1024^3  # Allow 4GB for data transfer
)

# 2. Process in batches by cell type
cell_types <- unique(large_seurat_obj$cell_type)
batch_results <- list()

for (ct in cell_types) {
  message(paste0("Processing cell type: ", ct))
  # Subset to this cell type
  ct_obj <- subset(large_seurat_obj, cell_type == ct)
  
  # Run analysis
  ct_result <- quickCompare(
    object = ct_obj,
    cell_types = ct,
    group1 = "Treatment",
    group2 = "Control"
  )
  
  # Store result
  batch_results[[ct]] <- ct_result
}

# 3. Combine batch results
all_degs <- do.call(rbind, batch_results)

# 4. Export to file for faster loading in future sessions
exportDEGResults(
  deg_results = all_degs,
  file_path = "large_dataset_degs.xlsx",
  split_by = "cell_type"
)
```

# Custom Analysis Workflows

CellHet functions can be combined to create custom analysis workflows:

```{r custom-workflow, eval=FALSE}
# Define a custom analysis workflow
multiConditionAnalysis <- function(seurat_obj, 
                                 cell_type_var = "cell_type",
                                 reference_condition = "Control") {
  
  # Get all conditions
  conditions <- unique(seurat_obj$condition)
  all_conditions <- setdiff(conditions, reference_condition)
  
  # Initialize results
  all_results <- list()
  
  # For each condition vs reference
  for (cond in all_conditions) {
    message(paste0("Analyzing ", cond, " vs ", reference_condition))
    
    # Subset to these two conditions
    subset_obj <- subset(seurat_obj, condition %in% c(cond, reference_condition))
    
    # Run DEG analysis
    deg_result <- findDifferentialGenes(
      object = subset_obj,
      group_var = "condition",
      cell_type_var = cell_type_var,
      reference_group = reference_condition
    )
    
    # Analyze heterogeneity
    het_result <- visualizeHeterogeneity(
      deg_results = deg_result,
      layout_type = "network"
    )
    
    # Run pathway analysis
    pathway_result <- runPathwayAnalysis(
      deg_results = deg_result
    )
    
    # Visualize pathways
    pathway_viz <- visualizePathways(
      pathway_results = pathway_result,
      plot_type = "dotplot"
    )
    
    # Store results
    all_results[[cond]] <- list(
      degs = deg_result,
      heterogeneity = het_result,
      pathways = pathway_result,
      plots = list(
        het_plot = het_result$plot,
        pathway_plot = pathway_viz
      )
    )
  }
  
  # Create comparison of all conditions
  response_comparison <- data.frame()
  
  for (cond in all_conditions) {
    # Extract summary data
    cond_summary <- all_results[[cond]]$degs$summary
    cond_summary$condition <- cond
    
    # Append to comparison
    response_comparison <- rbind(response_comparison, cond_summary)
  }
  
  # Create comparative barplot
  comp_plot <- ggplot(response_comparison, aes(x = cell_type, y = n_significant, fill = condition)) +
    geom_bar(stat = "identity", position = "dodge") +
    theme_minimal() +
    labs(title = "DEG Response Across Conditions",
         x = "Cell Type",
         y = "Number of DEGs") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  # Add to results
  all_results$comparison_plot <- comp_plot
  all_results$response_comparison <- response_comparison
  
  return(all_results)
}

# Example usage:
# multi_condition_results <- multiConditionAnalysis(seurat_obj)
# print(multi_condition_results$comparison_plot)
```

# Conclusion

CellHet provides a comprehensive toolkit for analyzing cell heterogeneity in single-cell RNA-seq data. With its flexible functions and intuitive visualization capabilities, you can quickly identify and interpret differential gene expression patterns across cell types and conditions.

For more information, refer to the package documentation or visit the GitHub repository at [https://github.com/guokai8/CellHet](https://github.com/guokai8/CellHet).
